Edge sign prediction in general graphs and Correlation Clustering

### 2017-08-25 cc/gtx_algorithm.tex
another failed proof of gtx number of iteration

We would like to prove a similar result on the edges so that the overall complexity of the
\gtx{} algorithm would be $O(|V_0|\log |V_0|+|E_0|)$. However, this does not follow directly from the
reduction of number of nodes. Indeed, we can have a situation where most of the edges of $E_t$ are
between the stars of $G_t$ and thus will remain in the next iteration. Still, we hope that
similarly to the previous proof, this high density of outer edges will lead to strong collapse in
the next iteration.

\iffalse
In the case no singleton are formed during one execution of \extractStar{}, the number of nodes is
reduced by at least a factor $2$ (because each star is made of at least two nodes). On the other hand, we can
construct a graph with singletons where the factor of reduction in number of nodes and edges can be
arbitrarily close to $1$. Consider \autoref{fig:gtx_noreduc}, where $c_1$ and $c_2$ have a degree
$p$ and their neighbors (in light gray) all have a degree $q$, which we can vary from $1$ to $p-1$.
At this step $t$, we thus have $|V_t| = 2(1+p)+pq$ and $|E_t| = 2(p+qp)$. Next we run \extractStar{}
and assume that we first extract the two stars centered in $c_1$ and $c_2$. This reduces the
effective degree of all the middle nodes (in dark gray) to $0$ and they thus form singletons, that
are all connected to the first two stars by \collapseStar{}. As a result, at step $t+1$ we have%
\begin{marginfigure}
  \centering
  \includegraphics[width=0.95\linewidth]{assets/tikz/gtx_noreduc_tikz.pdf}
  \caption{A case were too many singletons \enquote{waste} one iteration of \gtx{}}
  \label{fig:gtx_noreduc}
\end{marginfigure}
$|V_{t+1}| = 2 + pq$ and $|E_{t+1}| = 2pq$, so that
\begin{equation*}
\frac{|V_{t+1}|}{|V_t|} = \frac{pq+2}{p(q+2) + 2} \qquad \text{and} \qquad
\frac{|E_{t+1}|}{|E_t|} = \frac{2pq}{2p(q+1)} = \frac{q}{q+1}
\end{equation*}
When we let $p$ goes to infinity, setting $q=1$ makes $\frac{|V_{t+1}|}{|V_t|}$ tend to
$\nicefrac{1}{3}$ and $\frac{|E_{t+1}|}{|E_t|}$ to $\nicefrac{1}{2}$ while setting $q=p-1$ makes
both $\frac{|V_{t+1}|}{|V_t|}$ and $\frac{|E_{t+1}|}{|E_t|}$ tend to $1$. Note however that in both
cases, this is not really a problem because at the next iteration, there will be only two stars (one
of them will be a singleton). Is it always like that? In other words, can we have a situation where
a large number of singletons is carried over several consecutive iterations?
\fi

>>>>

Here is another proof that does not work, where we study the diameter of the graph. Indeed, it is
initially bounded by the number of nodes $n$, and by the correctness of \gtx{}, it will be zero upon
termination. If we show that the diameter shrinks by a constant factor at each iteration, then we
can claim that \gtx{} terminates after a number of iterations logarithmic in $n$. For ease of
exposition, we instead prove equivalently that when going backward in time, the diameter expands at
least by a constant factor at each iteration. Let us denote by $\Delta_t$ the diameter of $G_t$ and
take any path of length $\ell \leq \Delta_t$ in $G_t$. Let $[s_1, s_2, \ldots, s_{\ell+1}]$ be the
nodes along this path, where the notation $s_i$ reflects the fact that these nodes were stars of
nodes in $G_{t-1}$. If we choose any $u,v \in V_{t-1}^2$ such that $u$ is in the star $s_1$ and $v$
is in the star $s_{\ell+1}$, then we can see on \autoref{fig:gtx_expand_path} that there exists a
path of length $2\ell -1$ between $u$ and $v$ in $G_{t-1}$.
{\color{blue}{\large The figure actually shows this is not always true.}
More generally, for any path of length
$\ell$ in $G_t$, there is a path of length $2\ell-1$ in $G_{t-1}$. In particular, for
$\ell=\Delta_t$, there is a path of length $2\Delta_t-1$ in $G_{t-1}$, from which we conclude that
$2\Delta_t-1\leq\Delta_{t-1}$, or equivalently $\Delta_t \leq
\frac{1}{2}\left(\Delta_{t-1}+1\right)$. One can then check by recurrence that $\Delta_t \leq
\frac{1}{2^t} \left(\Delta_0+2^t-1\right) \leq \frac{n}{2^t} + 1$.}

\begin{figure*}[htpb]
   \centering
   \includegraphics[width=\linewidth]{raw/expand_path.pdf}
   \caption[Star expansion of a path]{A path within $G_t$ in black, and the underlying graph
     $G_{t-1}$ in blue. In the worst case, the endpoints of the path in $G_{t-1}$ are at the
     periphery of the endpoint stars.  The intermediary nodes in $G_t$ could be proper or singleton
     star, but there cannot be two adjacent singleton stars.}
   \label{fig:gtx_expand_path}
\end{figure*}

\iffalse
We would like to prove a similar result on the edges so that the overall complexity of the
\gtx{} algorithm would be $O(|V_0|\log |V_0|+|E_0|)$. However, this does not follow directly from the
reduction of number of nodes. Indeed, we can have a situation where most of the edges of $E_t$ are
between the stars of $G_t$ and thus will remain in the next iteration. Still, we hope that
similarly to the previous proof, this high density of outer edges will lead to strong collapse in
the next iteration.

In the case no singleton are formed during one execution of \extractStar{}, the number of nodes is
reduced by at least a factor $2$ (because each star is made of at least two nodes). On the other hand, we can
construct a graph with singletons where the factor of reduction in number of nodes and edges can be
arbitrarily close to $1$. Consider \autoref{fig:gtx_noreduc}, where $c_1$ and $c_2$ have a degree
$p$ and their neighbors (in light gray) all have a degree $q$, which we can vary from $1$ to $p-1$.
At this step $t$, we thus have $|V_t| = 2(1+p)+pq$ and $|E_t| = 2(p+qp)$. Next we run \extractStar{}
and assume that we first extract the two stars centered in $c_1$ and $c_2$. This reduces the
effective degree of all the middle nodes (in dark gray) to $0$ and they thus form singletons, that
are all connected to the first two stars by \collapseStar{}. As a result, at step $t+1$ we have%
\begin{marginfigure}
  \centering
  \includegraphics[width=0.95\linewidth]{assets/tikz/gtx_noreduc_tikz.pdf}
  \caption{A case were too many singletons \enquote{waste} one iteration of \gtx{}}
  \label{fig:gtx_noreduc}
\end{marginfigure}
$|V_{t+1}| = 2 + pq$ and $|E_{t+1}| = 2pq$, so that
\begin{equation*}
\frac{|V_{t+1}|}{|V_t|} = \frac{pq+2}{p(q+2) + 2} \qquad \text{and} \qquad
\frac{|E_{t+1}|}{|E_t|} = \frac{2pq}{2p(q+1)} = \frac{q}{q+1}
\end{equation*}
When we let $p$ goes to infinity, setting $q=1$ makes $\frac{|V_{t+1}|}{|V_t|}$ tend to
$\nicefrac{1}{3}$ and $\frac{|E_{t+1}|}{|E_t|}$ to $\nicefrac{1}{2}$ while setting $q=p-1$ makes
both $\frac{|V_{t+1}|}{|V_t|}$ and $\frac{|E_{t+1}|}{|E_t|}$ tend to $1$. Note however that in both
cases, this is not really a problem because at the next iteration, there will be only two stars (one
of them will be a singleton). Is it always like that? In other words, can we have a situation where
a large number of singletons is carried over several consecutive iterations?
\fi
